Modelling the Lymphatic Metastatic Progression Pathways of OPSCC from Multi-Institutional Datasets

### 2023-03-09 src/tex/ms.tex
[showyourwork] overleaf sync

% Define document class
\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}
\usepackage{multirow}
\usepackage{nicefrac}
\usepackage{wasysym}

@@

% Abstract
\begin{abstract}
    The \gls{ctv-n}\glsunset{ctv} definition in \gls{hnscc}\glsunset{scc} is currently based mostly on the prevalence of lymphatic metastases in different \glspl{lnl} based on the primary tumor location. In this work, we present an extension to a probabilistic model for lymphatic metastatic spread developed earlier that can quantify the risk for microscopic nodal involvement based on individual diagnoses. The extension is based on the same formalism of \glspl{hmm} as the original model, but in addition to the \glspl{lnl} I, II, III, and IV, it also covers the \glspl{lnl} V and VII of the ipsilateral neck. Moreover, we infer which pathways of lymphatic spread to model from clinical lymphatic progression patterns of 681 patients from three institutions. The extended model may allow for a more personalized \gls{ctv-n} definition based on a patient's individual state of disease. The \gls{hmm} uses a collection of hidden binary \glspl{rv} -- one for each \gls{lnl} -- to model a patient's state of lymphatic involvement. Clinical diagnoses and/or pathological examinations of resected \glspl{lnl} represent the observed binary \glspl{rv} corresponding to the unobservable true state of nodal disease. A \gls{dag} with parametrized edges is used to compute the transition matrix of the \gls{hmm} and consequently the (log-)likelihood of the diagnoses of 681 patients, given a set of spread parameters. Based on the defined likelihood function, \gls{mcmc} sampling is then used to infer a distribution over the model's parameters. T-category is model by assuming that, on average, patients with late T-category tumors are diagnosed at a later time compared to patients with early T-category tumors. Using \gls{ti} and the \gls{bic} we compare models based on different \glspl{dag} and demonstrate the accuracy and precision of the best-performing model. When extended to the contralateral side, this model may form the basis of future guidelines of elective \gls{ctv} definition.
\end{abstract}

@@

With the publication of larger and more detailed datasets \cite{ludwig_dataset_2022}, the risk estimation models may be extended as well to include the additionally reported \glspl{lnl} V and VII. This work aims to do that and at the same time demonstrate the capability 
of the previously introduced \gls{hmm} to adapt to real patient data from a multi-centric cohort of XXX \gls{opscc} patients.

\section{Previous Work}
\label{sec:previous_work}

\subsection{Lymphatic Spread as Hidden Markov Model}
\label{subsec:previous_work:spread_as_hmm}

We have introduced a probabilistic model for lymphatic metastatic tumor progression based on \glspl{bn} in \cite{pouymayou_bayesian_2019} and based off of this an improved model using \glspl{hmm} in \cite{ludwig_hidden_2021}. We will briefly recap the \acrlong{hmm} below.

A patient's state of (hidden) lymphatic involvement at time $t$ is described as a collection of binary \glspl{rv}, one for each of the $V$ \glspl{lnl}:
%
\begin{equation}
    \mathbf{X}[t] = \left( X_v[t] \right) \qquad v \in \left\{ 1,2, \ldots, V \right\}
\end{equation}
%
Where each of the \glspl{lnl} can be in the state $X_v=0$ (\texttt{FALSE}), meaning \gls{lnl} $v$ is healthy, or in the state $X_v=1$ (\texttt{TRUE}), indicating the \gls{lnl} harbors metastases.

The transition from one time-step to another is governed by the transition probability $P\left( \mathbf{X}[t+1]=\boldsymbol{\xi}_i \mid \mathbf{X}[t]=\boldsymbol{\xi}_j \right)$, which can conveniently be collected into a transition matrix when we enumerate all $2^V$ distinct possible states $\boldsymbol{\xi}_i$ with $i \in \left\{ 1,2, \ldots, 2^V \right\}$ of lymphatic involvement:
%
\begin{equation}
    \mathbf{A} = \left( A_{ij} \right) = \left( P\left( \mathbf{X}[t+1]=\boldsymbol{\xi}_i \mid \mathbf{X}[t]=\boldsymbol{\xi}_j \right) \right)
\end{equation}
%
The term $P\left( \boldsymbol{\xi}_i \mid \boldsymbol{\xi}_j \right)$ describes the probability to transition from the hidden state of lymphatic involvement $\boldsymbol{\xi}_j$ to the state $\boldsymbol{\xi}_i$ between the time $t$ and $t+1$. Using a \gls{dag} as depicted in \cref{fig:graph_with_obs}, we can parameterize this transition probability in the following way:
%
\begin{equation}
    \label{eq:transition_prob}
    P\left( \boldsymbol{\xi}_i \mid \boldsymbol{\xi}_j \right) = \prod_{v \leq V} Q\left( \xi_{iv} ; \xi_{jv} \right) P \left( \xi_{iv} \mid \left\{ \xi_{jr} \right\}_{r \in \operatorname{pa}(v)} \right)^{1 - \xi_{jv}}
\end{equation}
%
In this equation, we have denoted \glspl{lnl} that are parents of \gls{lnl} $v$ with the symbol $r\in\operatorname{pa}(v)$. Also, $\xi_{iv}$ denotes the value that \gls{lnl} $v$ takes on when the patient is in state $\boldsymbol{\xi}_i$. Lastly, the term $Q(a;b) \in \{ 0,1 \}$ is there to prohibit self-healing. It is always one, except if \gls{lnl} $v$ is healthy in state $\xi_i$, but was metastatic in the previous state $\xi_j$. In that case the function becomes $Q(0;1) = 0$, making the transition back to healthier states impossible.

\begin{figure}
    \centering
    \includegraphics[width=0.6\linewidth]{figures/graph_with_obs.png}
    \caption{\Gls{dag} representing a possible abstraction of the lymphatic network comprising the tumor (red shaded circle) and \glspl{lnl} I through IV as hidden binary \glspl{rv} (blue outlined circles). Attached to each of these is the corresponding observed \gls{rv} (orange shaded squares). Lymphatic flow is depicted in the form of parameterized arrows (red and blue) that represent the probability of spread along the respective arc per time-step. Sensitivity and specificity (orange arrows) connect the hidden \glspl{rv} to the diagnosis.}
    \label{fig:graph_with_obs}
\end{figure}

Diagnosis and true state of a patient are formally connected via the sensitivity $s_N$ and specificity $s_P$ of the used diagnostic modality. In clinical practice, these modalities are \gls{ct}, \gls{mri}, or \gls{pet} scan, but it may also include information from biopsies after a \gls{fna} or other techniques to detect lymphatic metastases. For each \gls{lnl} $v$ the conditional probability table of $P\left( Z_v \mid X_v \right)$ looks like this:

@@

If multiple diagnostic modalities $\mathcal{O} \in \{ \text{MRI}, \text{CT}, \ldots \}$ are used, one can compute an observation matrix $\mathbf{B}^\mathcal{O}$ for each of these modalities and combine them via an outer product into a joint observations matrix:
%
\begin{equation}
    \mathbf{B} = \mathbf{B}^\text{MRI} \otimes \mathbf{B}^\text{CT} \otimes \ldots
\end{equation}

We define the time $t=0$ to be the moment just before a patient's tumor formed, and hence $X_v[t=0]=0 \,\,\, \forall v$. However, using this definition, we cannot know how many time-steps have passed until $t_\text{D}$, when the patient was diagnosed with cancer. We can only make the assumption that a patients with an earlier T-category tumor was \emph{probably} diagnosed after fewer time-steps than a patient with a late T-category tumor. We can use this information by marginalizing over the diagnose times $t_\text{D}$ of patients in different T-categories using different prior distributions over the diagnose time. E.g., $P \left( t=t_D \mid \text{early} \right)$ for early T-category patients (T1 \& T2) and $P \left( t=t_D \mid \text{late} \right)$ for late T-category patients (T3 \& T4). Throughout this work we will use binomial distributions for these probability mass functions, mainly because they have a plausible shape for this purpose and only a single parameter.
%
\begin{equation}
    P \left( t = t_D \mid \text{T-cat.} \right) = \operatorname{B}(t_\text{max},p_\text{T-cat.})
\end{equation}
%
Here, the parameter $p_\text{early}$ can be interpreted as the probability that the patient will be diagnosed at time-step $t+1$ given they are in time-step $t$. We will use as a latest time-step $t_\text{max} = 10$.

\subsection{The Likelihood Function}
\label{subsec:previous_work:likelihood}

Using the definitions up to this point, we can compute a vector of likelihoods for every possible diagnosis:
%
\begin{equation} \label{eq:likelihood_vec}
\begin{split}
    \boldsymbol{\ell} &= \big( P\left( \mathbf{Z} = \boldsymbol{\zeta}_i \right) \big) \\
    &= \sum_{t=0}^{t_{max}} \left[ \boldsymbol{\pi} \cdot \mathbf{A}^t \cdot \mathbf{B} \right] \cdot P \left( t \mid \text{T-cat.} \right)
\end{split}
\end{equation}
%
This likelihood implicitly depends on how we parameterize the arcs of the \gls{dag} underlying the model and the parameterization of the distribution over diagnosis times. If we use the graph displayed in \cref{fig:graph_with_obs}, we can write down a conditional probability table for \gls{lnl} $X_3$ being healthy or metastatic, based on the state of the parent \gls{lnl} $\operatorname{pa}(3) = 2$:

Similar tables can be written for all other \glspl{lnl}. The terms in these conditional probability tables can then be used to compute the transition probability between to particular states $\boldsymbol{\xi}_i$ and $\boldsymbol{\xi}_j$ as in \cref{eq:transition_prob}. Computing this quantity for all possible transitions subsequently yields the transition matrix $\mathbf{A}$ as described in \cref{eq:transition_matrix}.

Together with the parametrizations of the distributions over the diagnosis time, the parameters that make up the transition matrix $\mathbf{A}$ comprises the set of model parameters:
%
\begin{equation}
    \boldsymbol{\theta} = \left( \left\{ b_v \right\}, \left\{ t_{vr} \right\}, p_\text{early}, p_\text{late} \right) \quad \text{with} \quad \genfrac{}{}{0pt}{2}{v\leq V}{r\in\operatorname{pa}(v)}
\end{equation}

@@

Using this log-likelihood function one may now employ a variety of inference methods to learn the parameters of the model that best describe the observed data. After that, the main task of of personalizing the \gls{ctv-n} definition is to predict the probability of the hidden possible states $\boldsymbol{\xi}_k$ given the diagnosis $d^\star=\boldsymbol{\zeta}_\ell$ of a new patient. Using Bayes' theorem, we get
%
\begin{equation}
    P\left( \mathbf{X}=\boldsymbol{\xi}_k \mid \mathbf{Z}=\boldsymbol{\zeta}_\ell \right) = \frac{P\left( \boldsymbol{\zeta}_\ell \mid \boldsymbol{\xi}_k \right) P\left( \boldsymbol{\xi}_k \mid \boldsymbol{\theta} \right)}{\sum_{r=1}^{2^V} P\left( \boldsymbol{\zeta}_\ell \mid \boldsymbol{\xi}_r \right) P\left( \boldsymbol{\xi}_r \mid \boldsymbol{\theta} \right) }
\end{equation}
%
The described model along with the inferred parameters $\boldsymbol{\hat{\theta}}$ will yield an estimate (or multiple estimates) for the ``prior'' in the above equation $P\big( \boldsymbol{\xi}_k \mid \boldsymbol{\hat{\theta}} \big)$.

\section{Methods}
\label{sec:methods}

\subsection{Parameter Inference}
\label{subsec:methods:inference}

@@

For the experiments in this work, we used $W = 20 \cdot k$ walkers, where $k$ is the dimensionaliry of the parameter space $\boldsymbol{\Theta}$. After an initial ``burn-in'' phase, during which all drawn samples are discarded because they are not yet independent from the initial state, we continued sampling for another 200 steps of which we discarded every 10th to be left with $S = 20 \cdot W$ samples.

@@

\subsection{Model Comparison}
\label{subsec:methods:comparison}

@@

The aim of this work is to refine the graph structure underlying our risk model introduced in the previous section. This \gls{dag} determines the number of parameters of the model as well as how exactly the transition matrix $\mathbf{A}$ is parameterized. To compare different models that are based on different \glspl{dag}, e.g. models $\mathcal{M}_1$ and $\mathcal{M}_2$, in a Bayesian setting, we need to compute the probabilities of these models, given the data $\boldsymbol{\mathcal{D}}$:
%
\begin{equation}
    P \left( \mathcal{M}_i \mid \boldsymbol{\mathcal{D}} \right) = \frac{P \left( \boldsymbol{\mathcal{D}} \mid \mathcal{M}_i \right) P \left( \mathcal{M}_i \right)}{ P \left( \boldsymbol{\mathcal{D}} \right) }
\end{equation}
%
If we assume all models $\mathcal{M}_i$ for $i \in \{ 1,2 \}$ to have the same \emph{a priori} probability -- meaning in this case $P (\mathcal{M}_1) = P (\mathcal{M}_2)$ -- then we can compute the so-called \emph{Bayes factor} of the two models as the ratio of their likelihoods. The interpretation of the values for different Bayes factors is given in \cref{table:bayes_factor}. It is defined as follows:
%
\begin{equation}
    K_\text{1v2} = \frac{P \left( \mathcal{M}_1 \mid \boldsymbol{\mathcal{D}} \right)}{P \left( \mathcal{M}_2 \mid \boldsymbol{\mathcal{D}} \right)} = \frac{P \left( \boldsymbol{\mathcal{D}} \mid \mathcal{M}_1 \right)}{P \left( \boldsymbol{\mathcal{D}} \mid \mathcal{M}_2 \right)}
\end{equation}
%
These likelihoods are commonly called the \emph{model evidence} or \emph{marginal likelihood}. The latter because computing it involves marginalizing the data likelihood over all model parameters:
%
\begin{equation} \label{eq:evidence}
    E_\mathcal{M} = P \left( \boldsymbol{\mathcal{D}} \mid \mathcal{M} \right) = \int_{\boldsymbol{\Theta}} P\left( \boldsymbol{\mathcal{D}} \mid \boldsymbol{\theta}, \mathcal{M} \right) p(\boldsymbol{\theta} \mid \mathcal{M}) d\boldsymbol{\theta}
\end{equation}
%
However, this quantity is often very hard to compute or even intractable, due to the high dimensionality of the parameter space $\boldsymbol{\Theta}$. In our case, the number of dimensions ranges from $k=9$ for the \emph{base graph} to $k=11$ for the \emph{winning graph}. A brute-force integration over a unit cube with this many dimensions is inefficient and error prone, which is why we resorted to \gls{ti} for computing the (log-)evidence.

@@

The derivatives of the log-evidences $\ln E_\beta$ are essentially expectation values of the data log-likelihood under the power posteriors of the corresponding value for $\beta$. They can be computed using \gls{mcmc}:
%
\begin{equation}
    \begin{aligned}
        \frac{d}{d\beta} \ln E_\beta &= \int p_\beta \left( \boldsymbol{\theta} \mid \boldsymbol{\mathcal{D}}, \mathcal{M} \right) \ln P \left( \boldsymbol{\mathcal{D}} \mid \boldsymbol{\theta}, \mathcal{M} \right) d\boldsymbol{\theta} \\
        &= \mathbb{E} \left[ \ln P \left( \boldsymbol{\mathcal{D}} \mid \boldsymbol{\theta}, \mathcal{M} \right) \right]_{ p_\beta \left( \boldsymbol{\theta} \mid \boldsymbol{\mathcal{D}}, \mathcal{M} \right) } \\
        &\approx \frac{1}{S} \sum_{i=1}^S \ln P \left( \boldsymbol{D} \mid \hat{\boldsymbol{\theta}}_{\beta i}, \mathcal{M} \right) =: \mathcal{A}_\text{MC} \left( \beta \right)
    \end{aligned}
\end{equation}
%
The integral in \cref{eq:ti} can then be computed via a trapezoidal rule using the $\mathcal{A}_\text{MC}$ to yield a  numerical approximation of the model evidence:
%
\begin{equation}
    \ln E \approx \frac{1}{2} \sum_{j=0}^{R-1} \left( \beta_{j+1} - \beta_j \right) \cdot \big( \mathcal{A}_\text{MC} (\beta_{j+1}) + \mathcal{A}_\text{MC} (\beta_j) \big)
\end{equation}
%
This estimate gets better for more samples $S$ per sampling from the power posterior $p_\beta$ but more importantly it gets better for a tighter spacing of the values for $\beta$ within the interval $[0,1]$. The variable $\beta$ is also often referred to as an \emph{inverse temperature}, due to its origins in statistical physics. Often when performing \gls{ti}, the most drastic changes in the values of the $\mathcal{A}_\text{MC}$ occur at high temperatures (meaning $\beta$ very close to zero), while the changes become smaller and smaller for lower temparatures ($\beta$ towards one). It is therefore efficient to space the \emph{temperature ladder} unevenly, e.g. according to a fifth order power rule:
%
\begin{equation} \label{eq:power_rule}
    \beta_j = \left( j / R \right)^5 \qquad j \in \{ 0, 1, 2, \ldots, R \}
\end{equation}
%
For the \glspl{ti} that were performed in this work we used such a fifth order power rule with 64 steps, meaning that $R=63$.

>>>>

% Define document class
\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}
\usepackage{multirow}
\usepackage{nicefrac}
\usepackage{wasysym}
\usepackage{amssymb}

@@

% Abstract
\begin{abstract}
    The \gls{ctv-n}\glsunset{ctv} in \gls{hnscc}\glsunset{scc} is currently based mostly on the prevalence of lymphatic metastases in different \glspl{lnl} for a given primary tumor location. In this work, we present an extension to a probabilistic model for lymphatic metastatic spread developed earlier that can quantify the risk for microscopic nodal involvement based on an individual patient diagnosis. The extension is based on the same formalism of \glspl{hmm} as the original model, but in addition to the \glspl{lnl} I, II, III, and IV, it also covers the \glspl{lnl} V and VII of the ipsilateral neck. Moreover, we infer which pathways of lymphatic spread to model from clinical lymphatic progression patterns of 681 patients from three institutions. The extended model may allow for a more personalized \gls{ctv-n} definition based on a patient's individual state of disease. The \gls{hmm} uses a collection of hidden binary \glspl{rv} -- one for each \gls{lnl} -- to model a patient's state of lymphatic involvement. Clinical diagnoses and/or pathological examinations of resected \glspl{lnl} represent the observed binary \glspl{rv} corresponding to the unobservable true state of nodal disease. A \gls{dag} with parametrized edges is used to compute the transition matrix of the \gls{hmm} and consequently the (log-)likelihood of the diagnoses of 681 patients, given a set of spread parameters. Based on the defined likelihood function, \gls{mcmc} sampling is then used to infer a distribution over the model's parameters. T-category is model by assuming that, on average, patients with late T-category tumors are diagnosed at a later time compared to patients with early T-category tumors. Using \gls{ti} and the \gls{bic} we compare models based on different \glspl{dag} and demonstrate the accuracy and precision of the best-performing model. When extended to the contralateral side, this model may form the basis of future guidelines of elective \gls{ctv} definition.
\end{abstract}

@@

With the publication of larger and more detailed datasets \cite{ludwig_dataset_2022}, we are thus able to
\begin{enumerate}
    \item include all relevant \glspl{lnl} into the model, namely the levels I, II, III, IV, V, and VII.
    \item train the probabilistic model based on a multi-centric dataset consisting of \variable{output/num_patients.txt} patients from %\variable{output/num_institutions.txt} institutions.
    \item present exaples of results on personalized risk estimations performed with the trained model.
\end{enumerate}

\section{HMM Formalism and Notation}
\label{sec:formalism}

\glsreset{bn}

\subsection{State of the Hidden Markov Model}
\label{subsec:formalism:state}

We have introduced a probabilistic model for lymph node involvement based on \glspl{bn} in \cite{pouymayou_bayesian_2019}. The model was extended using \glspl{hmm} in \cite{ludwig_hidden_2021}. We will briefly recap the \acrlong{hmm} to introduce the notation used throughout the work.

A patient's state of (hidden) lymphatic involvement at time $t$ is described as a collection of binary \glspl{rv}, one for each of the $V$ \glspl{lnl}:
%
\begin{equation}
    \mathbf{X}[t] = \left( X_v[t] \right) \qquad v \in \left\{ 1,2, \ldots, V \right\}
\end{equation}
%
Each the \gls{lnl} can be in the state $X_v=0$ (\texttt{FALSE}), meaning \gls{lnl} $v$ is healthy, or in the state $X_v=1$ (\texttt{TRUE}), indicating the \gls{lnl} harbors metastases. The involved state includes occult disease.

The transition from one time-step to another is governed by the transition probability $P\left( \mathbf{X}[t+1]=\boldsymbol{\xi}_i \mid \mathbf{X}[t]=\boldsymbol{\xi}_j \right)$, which can conveniently be collected into a transition matrix when we enumerate all $2^V$ distinct possible states $\boldsymbol{\xi}_i$ with $i \in \left\{ 1,2, \ldots, 2^V \right\}$ of lymphatic involvement:
%
\begin{equation}
    \mathbf{A} = \left( A_{ij} \right) = \left( P\left( \mathbf{X}[t+1]=\boldsymbol{\xi}_i \mid \mathbf{X}[t]=\boldsymbol{\xi}_j \right) \right)
\end{equation}
%
The term $P\left( \boldsymbol{\xi}_i \mid \boldsymbol{\xi}_j \right)$ describes the probability to transition from the hidden state of lymphatic involvement $\boldsymbol{\xi}_j$ to the state $\boldsymbol{\xi}_i$ between the time $t$ and $t+1$. Using a \gls{dag} as depicted in \cref{fig:graph_with_obs}, we can formulate this transition probability in the following way:
%
\begin{equation}
    \label{eq:transition_prob}
    P\left( \boldsymbol{\xi}_i \mid \boldsymbol{\xi}_j \right) = \prod_{v \leq V} Q\left( \xi_{iv} ; \xi_{jv} \right) P \left( \xi_{iv} \mid \left\{ \xi_{jr} \right\}_{r \in \operatorname{pa}(v)} \right)^{1 - \xi_{jv}}
\end{equation}
%
In this equation, we have denoted \glspl{lnl} that are parents of \gls{lnl} $v$ with the symbol $r\in\operatorname{pa}(v)$. Also, $\xi_{iv}$ denotes the value that \gls{lnl} $v$ takes on when the patient is in state $\boldsymbol{\xi}_i$. The term $Q(a;b) \in \{ 0,1 \}$ is there to prohibit self-healing. It is always one, except if \gls{lnl} $v$ is healthy in state $\xi_i$, but was metastatic in the previous state $\xi_j$. In that case the function becomes $Q(0;1) = 0$, making the transition back to healthier states impossible.

The terms of the form $P \left( \xi_{iv} \mid \left\{ \xi_{jr} \right\}_{r \in \operatorname{pa}(v)} \right)$ implicitly depend on how we parametrize e.g. the arcs of \cref{fig:graph_with_obs}. For example, if we look at the probability of spread to the \gls{lnl} $X_3$ depending on the state of that level's parent -- which, in this case, is $\operatorname{pa}(3) = 2$ -- we can write the different combinations into a conditional probability table as below.

The variable $b_3$ denotes the probability of lymphatic spread from the tumor to \gls{lnl} III during one time-step, and $t_{23}$ is the probability of spread from an involved level II further down the lymphatic chain into \gls{lnl} III.

\begin{figure}
    \centering
    \includegraphics[width=0.6\linewidth]{figures/graph_with_obs.png}
    \caption{\Gls{dag} representing a possible abstraction of the lymphatic network comprising the tumor (red shaded circle) and \glspl{lnl} II through IV as hidden binary \glspl{rv} (blue outlined circles). Attached to each of these is the corresponding observed \gls{rv} (orange shaded squares). Lymphatic flow is depicted in the form of parametrized arrows (red and blue) that represent the probability of spread along the respective arc per time-step. Sensitivity and specificity (orange arrows) connect the hidden \glspl{rv} to the diagnosis.}
    \label{fig:graph_with_obs}
\end{figure}

\subsection{Diagnostic Observation}
\label{subsec:formalism:diagnosis}

We also need to introduce a separate collection of \glspl{rv} that describe the diagnostic observation of a patient's involvement. In analogy to the hidden true state $\mathbf{X}[t]$ at time $t$, we write this diagnosis as
%
\begin{equation}
    \mathbf{Z} = \left( Z_v \right) \qquad v \in \left\{ 1,2, \ldots, V \right\}
\end{equation}
%
We do not need to differentiate between different times $t$ here, since a patient is ever only diagnosed once, after which treatment usually starts timely.
Diagnosis and true state of a patient are formally connected via the sensitivity $s_N$ and specificity $s_P$ of the used diagnostic modality. In clinical practice, these modalities are \gls{ct}, \gls{mri}, or \gls{pet} scan, but it may also include information from biopsies after a \gls{fna} or other techniques to detect lymphatic metastases. For each \gls{lnl} $v$ the conditional probability table of $P\left( Z_v \mid X_v \right)$ looks like this:

@@

We define the time $t=0$ to be the moment just before a patient's tumor formed, and hence $X_v[t=0]=0 \,\,\, \forall v$. However, using this definition, we cannot know how many time-steps have passed until $t_\text{D}$, when the patient was diagnosed with cancer. We can only make the assumption that a patients with an earlier T-category tumor was \emph{probably} diagnosed after fewer time-steps than a patient with a late T-category tumor. We can use this information by marginalizing over the diagnose times $t_\text{D}$ of patients in different T-categories using different prior distributions over the diagnose time. E.g., $P \left( t=t_D \mid \text{early} \right)$ for early T-category patients (T1 \& T2) and $P \left( t=t_D \mid \text{late} \right)$ for late T-category patients (T3 \& T4). Throughout this work we will use binomial distributions for these probability mass functions, mainly because they have a plausible shape for this purpose and only a single parameter.
%
\begin{equation} \label{eq:time_dist}
    P \left( t = t_D \mid \text{T}x \right) = \operatorname{\mathfrak{B}}(t_\text{max},p_{\text{T}x})
\end{equation}
%
Here, the parameter $p_{\text{T}x}$ can be interpreted as the probability that the patient with a tumor of T-category $x$ will be diagnosed at time-step $t+1$ given they are in time-step $t$. We will use as the latest time-step $t_\text{max} = 10$, which will therefore give us a distribution over the diagnosis time that has its mean at $\mathbb{E}[t_D] = 10 \cdot p_{\text{T}x}$.

\subsection{The Likelihood Function}
\label{subsec:formalism:likelihood}

Using the definitions up to this point, we can compute a vector of likelihoods for every possible diagnosis:
%
\begin{equation} \label{eq:likelihood_vec}
\begin{split}
    \boldsymbol{\ell} &= \big( P\left( \mathbf{Z} = \boldsymbol{\zeta}_i \right) \big) \\
    &= \sum_{t=0}^{t_{max}} \left[ \boldsymbol{\pi} \cdot \mathbf{A}^t \cdot \mathbf{B} \right] \cdot P \left( t \mid \text{T} \right)
\end{split}
\end{equation}
%
This likelihood implicitly depends on how we parameterize the arcs of the \gls{dag} underlying the model -- see \cref{eq:transition_prob} -- and the parameterization of the distribution over diagnosis times -- e.g., as in \cref{eq:time_dist}.

Together with the parametrizations of the distributions over the diagnosis time, the parameters $b_v$ and $t_{vr}$ that make up the transition matrix $\mathbf{A}$ comprises the set of model parameters:
%
\begin{equation}
    \boldsymbol{\theta} = \left( \left\{ b_v \right\}, \left\{ t_{vr} \right\}, p_\text{early}, p_\text{late} \right) \quad \text{with} \quad \genfrac{}{}{0pt}{2}{v\leq V}{r\in\operatorname{pa}(v)}
\end{equation}

@@

Using this log-likelihood function one may now employ a variety of inference methods to learn the parameters of the model that best describe the observed data.

\subsection{Parameter Inference}
\label{subsec:formalism:inference}

@@

For the experiments in this work, we used $W = 20 \cdot k$ walkers, where $k$ is the dimensionality of the parameter space $\boldsymbol{\Theta}$. After an initial ``burn-in'' phase, during which all drawn samples are discarded because they are not yet independent of the initial state, we continued sampling for another 200 steps of which we discarded every 10th to be left with $S = 20 \cdot W$ samples.

@@

\subsection{Risk estimation}
\label{subsec:formalism:risk}

Now, the main task of personalizing the \gls{ctv-n} definition is to predict the probability of the hidden possible states $\boldsymbol{\xi}_k$ given the diagnosis $d^\star=\boldsymbol{\zeta}_\ell$ of a new patient. Using Bayes' theorem, we get
%
\begin{equation}
    P\left( \mathbf{X}=\boldsymbol{\xi}_k \mid \mathbf{Z}=\boldsymbol{\zeta}_\ell \right) = \frac{P\left( \boldsymbol{\zeta}_\ell \mid \boldsymbol{\xi}_k \right) P\left( \boldsymbol{\xi}_k \mid \boldsymbol{\theta} \right)}{\sum_{r=1}^{2^V} P\left( \boldsymbol{\zeta}_\ell \mid \boldsymbol{\xi}_r \right) P\left( \boldsymbol{\xi}_r \mid \boldsymbol{\theta} \right) }
\end{equation}
%
The described model along with the inferred parameters $\boldsymbol{\hat{\theta}}$ will yield an estimate (or multiple estimates) for the ``prior'' in the above equation $P\big( \boldsymbol{\xi}_k \mid \boldsymbol{\hat{\theta}} \big)$.

\section{Complete Model of Ipsilateral Spread in OPSCC}
\label{sec:complete_model}

\subsection{Investigating Spread Graphs}
\label{subsec:complete_model:graphs}

The \gls{dag} shown in \cref{fig:graph_with_obs} includes the \glspl{lnl} II, III, and IV since previous models \cite{pouymayou_bayesian_2019,ludwig_hidden_2021} had to rely on a reconstructed dataset of \gls{opscc} patients that contained only information on \gls{lnl} I and these three levels \cite{sanguineti_defining_2009}. Now, with datasets available that almost always report the \glspl{lnl} I, II, III, IV, V, and VII, we may add more \glspl{rv} to extend the graph and investigate how these \glspl{rv} should be connected among each other.

We have more freedom in choosing how to connect the \glspl{lnl} to each other. To investigate which connections to add we start by establishing a baseline from a model using a minimal \emph{base graph}. It contains only the connections from \gls{lnl} II to III and an arc from \gls{lnl} III to IV, as motivated by the \emph{main lymphatic pathway} \cite{lengele_anatomical_2007}. In contrast to the \gls{dag} used in \cite{pouymayou_bayesian_2019,ludwig_hidden_2021}, we decided not to include the edge from level I to II into the base graph, since \gls{lnl} I is rarely involved and inferring $t_{12}$ is therefore difficult. Rather, we wanted to use the model comparison methodology outlined in \cref{subsec:complete_model:comparison} to see if the arc from \gls{lnl} I to II can be motivated from our data or if the arc should be flipped and go from \gls{lnl} II to I.

\subsection{Model Comparison}
\label{subsec:complete_model:comparison}

@@

The aim of this work is to refine the graph structure underlying our risk model introduced in the previous section. This \gls{dag} determines the number of parameters of the model as well as how exactly the transition matrix $\mathbf{A}$ is parameterized. To compare different models that are based on different \glspl{dag}, e.g. models $\mathcal{M}_1$ and $\mathcal{M}_2$, in a Bayesian setting, we need to compute the probabilities of these models, given the data $\boldsymbol{\mathcal{D}}$:
%
\begin{equation}
    P \left( \mathcal{M}_i \mid \boldsymbol{\mathcal{D}} \right) = \frac{P \left( \boldsymbol{\mathcal{D}} \mid \mathcal{M}_i \right) P \left( \mathcal{M}_i \right)}{ P \left( \boldsymbol{\mathcal{D}} \right) }
\end{equation}
%
If we assume all models $\mathcal{M}_i$ for $i \in \{ 1,2 \}$ to have the same \emph{a priori} probability -- meaning in this case $P (\mathcal{M}_1) = P (\mathcal{M}_2)$ -- then we can compute the so-called \emph{Bayes factor} of the two models as the ratio of their likelihoods. The interpretation of the values for different Bayes factors is given in \cref{table:bayes_factor}. It is defined as follows:
%
\begin{equation}
    K_\text{1v2} = \frac{P \left( \mathcal{M}_1 \mid \boldsymbol{\mathcal{D}} \right)}{P \left( \mathcal{M}_2 \mid \boldsymbol{\mathcal{D}} \right)} = \frac{P \left( \boldsymbol{\mathcal{D}} \mid \mathcal{M}_1 \right)}{P \left( \boldsymbol{\mathcal{D}} \mid \mathcal{M}_2 \right)}
\end{equation}
%
These likelihoods are commonly called the \emph{model evidence} or \emph{marginal likelihood}. The latter because computing it involves marginalizing the data likelihood over all model parameters:
%
\begin{equation} \label{eq:evidence}
    E_\mathcal{M} = P \left( \boldsymbol{\mathcal{D}} \mid \mathcal{M} \right) = \int_{\boldsymbol{\Theta}} P\left( \boldsymbol{\mathcal{D}} \mid \boldsymbol{\theta}, \mathcal{M} \right) p(\boldsymbol{\theta} \mid \mathcal{M}) d\boldsymbol{\theta}
\end{equation}
%
However, this quantity is often very hard to compute or even intractable, due to the high dimensionality of the parameter space $\boldsymbol{\Theta}$. In our case, the number of dimensions ranges from $k=9$ for the \emph{base graph} to $k=11$ for the \emph{winning graph}. A brute-force integration over a unit cube with this many dimensions is inefficient and error-prone, which is why we resorted to \gls{ti} for computing the (log-)evidence.

@@

The derivatives of the log-evidences $\ln E_\beta$ are essentially expectation values of the data log-likelihood under the power posteriors of the corresponding value for $\beta$. They can be computed using \gls{mcmc}:
%
\begin{equation}
    \begin{aligned}
        \frac{d}{d\beta} \ln E_\beta &= \int p_\beta \left( \boldsymbol{\theta} \mid \boldsymbol{\mathcal{D}}, \mathcal{M} \right) \ln P \left( \boldsymbol{\mathcal{D}} \mid \boldsymbol{\theta}, \mathcal{M} \right) d\boldsymbol{\theta} \\
        &= \mathbb{E} \left[ \ln P \left( \boldsymbol{\mathcal{D}} \mid \boldsymbol{\theta}, \mathcal{M} \right) \right]_{ p_\beta \left( \boldsymbol{\theta} \mid \boldsymbol{\mathcal{D}}, \mathcal{M} \right) } \\
        &\approx \frac{1}{S} \sum_{i=1}^S \ln P \left( \boldsymbol{D} \mid \hat{\boldsymbol{\theta}}_{\beta i}, \mathcal{M} \right) =: \mathcal{A}_\text{MC} \left( \beta \right)
    \end{aligned}
\end{equation}
%
The integral in \cref{eq:ti} can then be computed via a trapezoidal rule using the $\mathcal{A}_\text{MC}$ to yield a  numerical approximation of the model evidence:
%
\begin{equation}
    \ln E \approx \frac{1}{2} \sum_{j=0}^{R-1} \left( \beta_{j+1} - \beta_j \right) \cdot \big( \mathcal{A}_\text{MC} (\beta_{j+1}) + \mathcal{A}_\text{MC} (\beta_j) \big)
\end{equation}
%
This estimate gets better for more samples $S$ per sampling from the power posterior $p_\beta$ but more importantly it gets better for a tighter spacing of the values for $\beta$ within the interval $[0,1]$. The variable $\beta$ is also often referred to as an \emph{inverse temperature}, due to its origins in statistical physics. Often when performing \gls{ti}, the most drastic changes in the values of the $\mathcal{A}_\text{MC}$ occur at high temperatures (meaning $\beta$ very close to zero), while the changes become smaller and smaller for lower temperatures ($\beta$ towards one). It is therefore efficient to space the \emph{temperature ladder} unevenly, e.g. according to a fifth order power rule:
%
\begin{equation} \label{eq:power_rule}
    \beta_j = \left( j / R \right)^5 \qquad j \in \{ 0, 1, 2, \ldots, R \}
\end{equation}
%
For the \glspl{ti} that were performed in this work we used such a fifth order power rule with 64 steps, meaning that $R=63$.
